Map{R}educe in {MPI} for Large-scale Graph Algorithms

### 2010-06-07 paper/intro.tex
git-svn-id: svn+ssh://software-srn.sandia.gov/svn/mapreduce/trunk@630 3fc257bb-8550-0410-81b3-a5c103da83f7

\item {\it C++ library using MPI for inter-processor communication:}
The user writes a (typically) simple main program which runs on each
processor of a parallel machine, making calls to the MR-MPI library.
For map and reduce operations, the library calls back to user-provided
{\it map()} and {\it reduce()} functions.  The use of C++ allows precise control
over the memory and format of data allocated by each processor during
a MapReduce.  Library calls for performing a map, reduce, or data
shuffle, are synchronous, meaning all the processors participate and
finish the operation before proceeding.  Similarly, the use of MPI
within the library is the traditional mode of MPI\_Send and
MPI\_Recv calls between processor pairs using large aggregated
messages to improve bandwidth performance and reduce latency costs.  A
recent paper~\cite{Hoefler09} also outlines the MapReduce formalism
from an MPI perspective, although they advocate a more asynchronous
approach, using one-way communication of small messages.

@@

\item {\it In-core or out-of-core operation:} Each MapReduce object that a
processor defines allocates per-processor ``pages'' of memory, where
the page size is determined by the user.  Typical MapReduce operations
can be performed using a few such pages.  If the data set fits in a
single page (per processor), the library performs its operations
in-core.  If the data set exceeds the page size, processors each
write to temporary disk files (on local disks or a parallel file
system) as needed and subsequently read from them.  This out-of-core
operation allows
processing of data sets larger than the aggregate memory of all the
processors, i.e. up to the available aggregate disk space.

\item {\it Flexible programmability:} An advantage of writing a
MapReduce program on top of MPI, is that the user program can invoke
MPI calls directly, if desired.  For example, one-line calls to
MPI\_Allreduce are often useful in determining the status of an
iterative graph algorithm, as described in Section~\ref{sec:graph}.
The library interface also provides a user data pointer as an argument
passed back to all callback functions, so it is easy for the user
program to store ``state'' on each processor, accessible during the
map and reduce operations.  For example, various flags can be stored
that alter the operation of a map or reduce operation, as can richer
data structures that accumulate results.

@@

The remainder of the paper is organized as follows.  
Sections~\ref{sec:mr} and~\ref{sec:outcore} describe how in-core and
out-of-core MapReduce primitives are formulated as MPI-based
operations in the MR-MPI library.  Section~\ref{sec:graph} briefly
describes the formulation of several common graph algorithms as
MapReduce operations.  Section~\ref{sec:results} gives performance
results for these algorithms running on a parallel cluster for graphs
%KDDKDD CHECK SIZES
ranging in size from 8 million to 2 billion edges;  
we highlight the performance and complexity trade-offs
of a MapReduce approach versus other more special-purpose algorithms.
The latter generally perform better but are harder to implement
efficiently on distributed memory machines, due to the required
explicit management of parallelism, particularly for large out-of-core
data sets.  Section~\ref{sec:lessons} summarizes some lessons learned
from the implementation and use of our library.

>>>>

\item {\it C++ library using MPI for inter-processor communication:}
The user writes a (typically) simple main program which runs on each
processor of a parallel machine, making calls to the MR-MPI library.
For map and reduce operations, the library calls back to user-provided
{\it map()} and {\it reduce()} functions.  The use of C++ allows
precise control over the memory and format of data allocated by each
processor during a MapReduce.  Library calls for performing a map,
reduce, or data shuffle, are synchronous, meaning all the processors
participate and finish the operation before proceeding.  Similarly,
the use of MPI within the library is the traditional mode of MPI\_Send
and MPI\_Recv calls between pairs of processors using large aggregated
messages to improve bandwidth performance and reduce latency costs.  A
recent paper~\cite{Hoefler09} also outlines the MapReduce formalism
from an MPI perspective, although they advocate a more asynchronous
approach, using one-way communication of small messages.

@@

\item {\it In-core or out-of-core operation:} Each MapReduce object
that a processor defines allocates ``pages'' of memory, where the page
size is determined by the user.  Typical MapReduce operations can be
performed using a few such pages (per processor).  If the data set
fits in a single page (per processor), the library performs its
operations in-core.  If the data set exceeds the page size, processors
each write to temporary disk files (on local disks or a parallel file
system) as needed and subsequently read from them.  This out-of-core
operation allows processing of data sets larger than the aggregate
memory of all the processors, i.e. up to the available aggregate disk
space.

\item {\it Flexible programmability:} An advantage of writing a
MapReduce program on top of MPI, is that the user program can invoke
MPI calls directly, if desired.  For example, one-line calls to
MPI\_Allreduce are often useful in determining the status of an
iterative graph algorithm, as described in Section~\ref{sec:graph}.
The library interface also provides a user data pointer as an argument
passed to all callback functions, so it is easy for the user program
to store ``state'' on each processor, accessible during the map and
reduce operations.  For example, various flags can be stored that
alter the operation of a map or reduce operation, as can richer data
structures that accumulate results.

@@

The remainder of the paper is organized as follows.
Sections~\ref{sec:mr} and~\ref{sec:outcore} describe how in-core and
out-of-core MapReduce primitives are formulated as MPI-based
operations in the MR-MPI library.  Section~\ref{sec:graph} briefly
describes the formulation of several common graph algorithms as
MapReduce operations.  Section~\ref{sec:results} gives performance
results for these algorithms running on a parallel cluster for graphs
ranging in size from 8 million to 2 billion edges; we highlight the
performance and complexity trade-offs of a MapReduce approach versus
other more special-purpose algorithms.  The latter generally perform
better but are harder to implement efficiently on distributed memory
machines, due to the required explicit management of parallelism,
particularly for large out-of-core data sets.
Section~\ref{sec:lessons} summarizes some lessons learned from the
implementation and use of our library.

### 2010-06-07 paper/lessons.tex
git-svn-id: svn+ssh://software-srn.sandia.gov/svn/mapreduce/trunk@630 3fc257bb-8550-0410-81b3-a5c103da83f7

MapReduce achieves parallelism through randomizing the distribution of
data across processors, which often intentionally ignores data
locality.  This translates into maximal data movement (during a
shuffle) with communication between all pairs of processors.  But the
benefit is often good load-balance, even for hard-to-balance irregular
data sets.  By contrast, more traditional distributed-memory parallel
algorithms, e.g. for matrix operations, or grid-based or
particle-based simulation codes, tend to work hard to localize data
and minimize communication.  To do this they typically require a lot
of application-specific logic and parallel communication coding to
create and maintain a data decomposition, generate ghost versions of
nearby spatially-decomposed data, etc.

MapReduce algorithms can be hard to design, but are often relatively
easy to write and debug.  Thinking about a computational task from a
MapReduce perspective is different from traditional distributed-memory
parallel algorithm design.  For example, with an MPI mindset, it often
seems heretical to intentionally ignore data locality.  However,
writing small {\it map()} and {\it reduce()} functions is typically easy.  And
writing an algorithm that involves complex parallel operations
without actually needing to write application-specific parallel code
to move data via MPI calls is often a pleasant surprise.  Moreover,
if the MapReduce algorithm is initially coded so that it runs
correctly on one processor, it often works out-of-the-box on hundreds
or thousands of processors, without the need for additional debugging.

Performing MapReduce operations on a fixed allocation of processors on
a traditional MPI-based parallel machine is a somewhat different
conceptual model than that of cloud-computing MapReduces using (for
example) Hadoop.  In the former case, one can potentially control
which processor owns which data at various stages of an algorithm.
This is somewhat hidden from the user in a typical cloud-computing
model, where data simply exists somewhere in the cloud and Hadoop
ensures data moves where it is needed and is operated on by some
processor.  The cloud model is a nice data-centric abstraction which
allows for fault tolerance both to data loss (via redundant storage)
and to processor failure (via reassignment of work), neither of which
is typically possible on current MPI-based parallel machines.

However, since the MPI implementation of MapReduce as
described in this paper is processor-centric, one can sometimes
fruitfully exploit the possibility for processors to maintain
``state'' over the course of multiple map and reduce operations.  By
controlling where data resides for maps and reduces (e.g. via a
user-specified hash function), and by assuming that processor will
always be available, more efficient operations with less data movement
are sometimes possible.  The discussion of enhanced graph algorithms
in Section \ref{sec:graph} illustrated this.  To fully exploit this
idea for large data sets, mechanisms are needed for processors to
store, retrieve, and efficiently find needed datums on local disk
(e.g. static edges of a graph), so that archived state can be used
efficiently during subsequent map and reduce operations.

Finally, though this paper focuses on graph algorithms expressed as
MapReduce operations, there is nothing about the MR-MPI library itself
that is graph centric.  We hope the library can be generally useful on
large-scale monolithic or cloud-style parallel machines that support
MPI, for a variety of data-intense or compute-intense problems that
are amenable to solution using a MapReduce paradigm.

>>>>

MapReduce achieves parallelism through randomizing the distribution of
data across processors, which often intentionally ignores data
locality.  This translates into maximal data movement (during a data
shuffle or {\it collate()} operation) with communication between all
pairs of processors.  But the benefit is often good load-balance, even
for hard-to-balance irregular data sets.  By contrast, more
traditional distributed-memory parallel algorithms, e.g. for matrix
operations, or grid-based or particle-based simulation codes,
typically work hard to localize data and minimize communication.  To
do this they often require a lot of application-specific logic and
parallel communication coding to create and maintain a data
decomposition, generate ghost versions of nearby spatially-decomposed
data, etc.

MapReduce algorithms can be hard to design, but are often relatively
easy to write and debug.  Thinking about a computational task from a
MapReduce perspective is different from traditional distributed-memory
parallel algorithm design.  For example, with an MPI mindset, it often
seems heretical to intentionally ignore data locality.  However,
writing small {\it map()} and {\it reduce()} functions is typically
easy.  And writing an algorithm that involves complex parallel
operations without needing to write application-specific code to
communicate data via MPI calls or to move data between out-of-core
disk files and processor memory is often a pleasant surprise.
Moreover, if the MapReduce algorithm is initially coded so that it
runs correctly on one processor, it often works out-of-the-box on
hundreds or thousands of processors, without the need for additional
debugging.

Performing MapReduce operations on a fixed allocation of processors on
a traditional MPI-based parallel machine is a somewhat different
conceptual model than that of performing MapReduce operations on a
cloud computer using (for example) Hadoop.  In the former case, one
can potentially control which processor owns which data at various
stages of an algorithm.  This is somewhat hidden from the user in a
typical cloud-computing model, where data simply exists somewhere in
the cloud and Hadoop ensures data moves where it is needed and is
operated on by some processor.  The cloud model is a nice data-centric
abstraction which allows for fault tolerance both to data loss (via
redundant storage) and to processor failure (via reassignment of
work), neither of which is typically possible on current MPI-based
parallel machines.

However, since the MPI implementation of MapReduce as described in
this paper is processor-centric, one can sometimes fruitfully exploit
the possibility for processors to maintain ``state'' over the course
of multiple map and reduce operations.  By controlling where data
resides for maps and reduces (e.g. via a user-specified hash
function), and by assuming that processor will always be available,
more efficient operations with less data movement are sometimes
possible.  The discussion of enhanced graph algorithms in Section
\ref{sec:graph} illustrated this.  To fully exploit this idea for
large data sets, mechanisms and data structures are needed for
processors to store, retrieve, and efficiently find needed datums on
local disk (e.g. static edges of a graph), so that archived state can
be used efficiently during subsequent map and reduce operations.

Finally, though this paper focuses on graph algorithms expressed as
MapReduce operations, there is nothing about the MR-MPI library itself
that is graph specific.  We hope the library can be generally useful
on large-scale monolithic or cloud-style parallel machines that
support MPI, for a variety of data-intensive or compute-intensive
problems that are amenable to solution using a MapReduce paradigm.

### 2010-06-07 paper/mr.tex
git-svn-id: svn+ssh://software-srn.sandia.gov/svn/mapreduce/trunk@630 3fc257bb-8550-0410-81b3-a5c103da83f7

The basic datums stored and operated on by any MapReduce framweork are
key/value (KV) pairs.  In the MR-MPI library, individual keys or
values can be of any data type or length, or combinations of multiple
types (one integer, a string of characters, two integers and a double,
etc); they are simply treated as byte strings by the library.  A KV
pair always has a key; a KV's value may be NULL.  A related data type is
the key/multivalue (KMV) pair, where all values associated with the
same key are collected and stored contiguously as a multivalue, which
is just a longer byte string with an associated vector of integer
lengths, one per value.

A typical MR-MPI program makes at least three calls to the MR-MPI library,
to perform a {\it map()}, {\it collate()}, and {\it reduce()}
operation.  In a {\it map()}, zero or more key/value pairs are
generated by each processor.  Typically, this is done using data read
from files, but a {\it map()} may generate data itself or process
existing KV pairs to create new ones.  The KV pairs produced are
stored locally by each processor; a {\it map()} thus requires no
inter-processor communication.  Users call the library with a count of
tasks to perform and a pointer to a user function; the MR-MPI {\it
map()} operation invokes the user function multiple times as a
callback.  Depending on which variant of {\it map()} is called, the
user function may be passed a file name, a chunk of bytes from a large
file, a task ID, or a KV pair.  Options for assigning map tasks to processors
are specified by the user and include assigning consecutive 
chunks of tasks to processors, striding the tasks across processors,
or using a master-slave model that is useful when tasks have widely
varying workloads.

The {\it collate()} operation (or data shuffle in Hadoop) identifies
unique keys and collects all the values associated with those keys to
create KMV pairs.  This is done in two stages, the first of which
requires communication, since KV pairs with the same key may be owned
by any processor.  Each processor hashes each of its keys to determine
which processor will ``own'' it.  The $k$-byte length key is hashed into
a 32-bit value whose remainder modulo $P$ processors generates the owning
processor rank.  Alternatively, the user can provide a hash function
which converts the key into a processor rank.  Each processor then sends
each of its KV pairs to the owning processor.

After receiving new KV pairs, the second stage is an on-processor
computation, requiring no further communication.  Each processor
reorganizes its KV pairs into KMV pairs, one for each unique key it
owns.  This is done using a hash table, rather than a sort.  A sort
requires $\log(N)$ passes through the $N$ KV pairs.  With hashing,
the list of KMV
pairs can be created in two passes through the KV pairs. The first 
pass populates
the hash table with needed count and length information; the
second pass copies the key and value datums into the appropriate
location in a new KMV data structure.  Since the lookup of a key in a
well-formed hash table is a constant-time $O(1)$ operation, the cost
of the data reorganization is $O(N)$, rather than $O(N\log(N))$ for full
sorting.  It also has the added benefit that the number of values in each
KMV pair is known and is passed to the user function during a
reduce.  For some reduce operations, this count is all the information a
reduce requires; thus the values need not be looped over.  This feature is not
available in Hadoop-style data reorganizations, which sort the values; 
with sorting, the number of values associated with the same key is not known 
without interating over the values.

Note that the first portion of the {\it collate()} operation involves
all-to-all communication (each processor sends and receives data from
every other processor) driven by a distributed hash table.  The
communication can either be done via a MPI\_Alltoall() library call,
or by a custom routine that aggregates messages and invokes pointwise
MPI\_Send() and MPI\_IRecv() calls.

@@

Several related MapReduce operations are provided by the library and
exposed in the library interface.  For example, the {\it collate()} function
described above
calls two other functions:  {\it aggregate()}, which performs the all-to-all
communication, and {\it convert()}, which turns a list of KV pairs into
KMV pairs.  
Both {\it aggregate()} and {\it convert()} are exposed in the MR-MPI
interface for direct access by the user.
The {\it compress()} function allows on-processor operations to be 
performed; it is equivalent to a {\it convert()} with on-processor KVs as
input, followed by a {\it reduce()}.  
The {\it clone()} operation turns a list of KV pairs into
KMV pairs, with one value per key.  The {\it collapse()} operation turns
$N$ KV pairs into one KMV pair, with the keys and values of the KV
pairs becoming $2N$ values of a single multivalue assigned to a new
key on the KMV.  The {\it gather()} operation collects KV pairs from
all processors to a subset of processors; it is useful for doing
output from a single or subset of processors.  Library calls for
sorting datums by key or value or within multivalues are also
provided.  These routines invoke the C-library {\it quicksort()} function to
compute the sorted ordering of the KV pairs (or values within a
multivalue), using a user-provided comparison function.  The KV pairs
(or values in a multivalue) are then copied one-by-one into a new data
structure in sorted order.

@@

The above discussion assumed that the KV or KMV pairs stored by a
processor fit in its physical memory.  In this case, MR-MPI 
performs only ``in-core'' processing and no disk files are written
or read by any of the processors, aside from initial input data if it
exists or final output data if it is generated.  Note that the
aggregate physical memory of large parallel machines can be multiple
terabytes, which allows for large data sets to be processed in-core,
assuming the KV and KMV pairs remain evenly distributed across
processors throughout the sequence of MapReduce operations.  The use
of hashing to assign keys to processors is designed for
such load-balancing.  In the next section, we discuss what happens 
when data sets do not fit in available memory.

>>>>

The basic datums stored and operated on by any MapReduce framweork are
key/value (KV) pairs.  In the MR-MPI library, individual keys or
values can be of any data type or length, or combinations of multiple
types (one integer, a string of characters, two integers and a double,
etc); they are simply treated as byte strings by the library.  A KV
pair always has a key; a KV's value may be NULL.  A related data type
is the key/multivalue (KMV) pair, where all values associated with the
same key are collected and stored contiguously as a multivalue, which
is just a longer byte string with an associated vector of lengths, one
integer length per value.  In this section, we assume the datums
operated on by each processor fit in local memory.

A typical MR-MPI program makes at least three calls to the MR-MPI
library, to perform {\it map()}, {\it collate()}, and {\it reduce()}
operations.  In a {\it map()}, zero or more key/value pairs are
generated by each processor.  Often this is done using data read from
files, but a {\it map()} may generate data itself or process existing
KV pairs to create new ones.  The KV pairs produced are stored locally
by each processor; a {\it map()} thus requires no inter-processor
communication.  Users call the library with a count of tasks to
perform and a pointer to a user function; the MR-MPI {\it map()}
operation invokes the user function multiple times as a callback.
Depending on which variant of {\it map()} is called, the user function
may be passed a file name, a chunk of bytes from a large file, a task
ID, or a KV pair.  Options for assigning map tasks to processors are
specified by the user and include assigning consecutive chunks of
tasks to processors, striding the tasks across processors, or using a
master-slave model that is useful when tasks have widely varying
workloads.

The {\it collate()} operation (or data shuffle in Hadoop) identifies
unique keys and collects all the values associated with those keys to
create KMV pairs.  This is done in two stages, the first of which
requires communication, since KV pairs with the same key may be owned
by any processor.  Each processor hashes each of its keys to determine
which processor will ``own'' it.  The $k$-byte length key is hashed
into a 32-bit value whose remainder modulo $P$ processors is the
owning processor rank.  Alternatively, the user can provide a hash
function which converts the key into a processor rank.  Each processor
then sends each of its KV pairs to the owning processor.

After receiving new KV pairs, the second stage is an on-processor
computation, requiring no further communication.  Each processor
reorganizes its KV pairs into KMV pairs, one for each unique key it
owns.  This is done using a hash table, rather than a sort.  A sort
scales as $N\log_2(N)$, i.e. it requires $\log_2(N)$ passes through
the $N$ KV pairs.  With hashing, the list of KMV pairs can be created
in two passes.  The first pass populates the hash table with needed
count and length information; the second pass copies the key and value
datums into the appropriate location in a new KMV data structure.
Since the cost to lookup a key in a well-formed hash table is a
constant-time $O(1)$ operation, the cost of the data reorganization is
also $O(N)$.  This methodology has the added benefit that the number
of values in each KMV pair is known and can be passed to the user
function during a reduce.  For some reduce operations, this count is
all the information a reduce requires; the values need not be looped
over.  The count is not available in Hadoop-style data shuffles, which
sort the values; counting the values associated with the same key
requires iterating over the values.

Note that the first portion of the {\it collate()} operation involves
all-to-all communication (each processor sends and receives data from
every other processor) using a distributed hash table.  The
communication can either be done via a MPI\_Alltoall() library call,
or by a custom routine that aggregates messages and invokes
point-to-point MPI\_Send() and MPI\_IRecv() calls.

@@

Several related MapReduce operations are provided by the library.  For
example, the {\it collate()} function described above calls two other
functions: {\it aggregate()}, which performs the all-to-all
communication, and {\it convert()}, which turns a list of KV pairs
into KMV pairs.  Both functions can be called directly.  The {\it
compress()} function allows on-processor operations to be performed;
it is equivalent to a {\it convert()} with on-processor KVs as input,
followed by a {\it reduce()}.  The {\it clone()} function turns a list
of KV pairs into KMV pairs, with one value per key.  The {\it
collapse()} function turns $N$ KV pairs into one KMV pair, with the
keys and values of the KV pairs becoming $2N$ values of a single
multivalue assigned to a new key.  The {\it gather()} function
collects KV pairs from all processors to a subset of processors; it is
useful for doing output from one or a few processors.  Library calls
for sorting datums by key or value or for sorting the values within
each multivalue are also provided.  These routines invoke the
C-library {\it quicksort()} function to compute the sorted ordering of
the KV pairs (or values within a multivalue), using a user-provided
comparison function.  The KV pairs (or values in a multivalue) are
then copied into a new data structure in sorted order.

@@

The above discussion assumed that the KV or KMV pairs stored by a
processor fit in its physical memory.  In this case, MR-MPI performs
only ``in-core'' processing and no disk files are written or read by
any of the processors, aside from initial input data if it exists or
final output data if it is generated.  Note that the aggregate
physical memory of large parallel machines can be multiple terabytes,
which allows for large data sets to be processed in-core, assuming the
KV and KMV pairs remain evenly distributed across processors
throughout the sequence of MapReduce operations.  The use of hashing
to assign keys to processors typically provides for such
load-balancing.  In the next section, we discuss what happens when
data sets do not fit in available memory.

### 2010-06-07 paper/outcore.tex
git-svn-id: svn+ssh://software-srn.sandia.gov/svn/mapreduce/trunk@630 3fc257bb-8550-0410-81b3-a5c103da83f7

When a user program creates a MapReduce object, a ``pagesize'' can be
specified, which defaults to 64 Mbytes.  As described below, each
MR-MPI operation is constrained to use no more than a handful of
these pages.  The {\it pagesize} setting can be as small as 1 Mbyte or
as large as desired, though the user should ensure the allocated pages
fit in physical memory; otherwise, the MR-MPI library may allocate slow
virtual memory.  The {\it pagesize} is also the typical size of
individual reads and writes to the temporary disk files; hence a
reasonable {\it pagesize} ensures good I/O performance.

@@

When the communication stage of the {\it collate()} operation is
complete, each processor has a set of KV pages, stored in an
out-of-core KV file, that need to be reorganized into a set of KMV
pages, stored in a new out-of-core KMV file.  In principle, 
creating one page of KMV pairs would require all KV pages be scanned
to find all the keys that contribute values to that page.  Doing this
for each output page of KMV pairs could be prohibitively expensive.  A
related issue is that, as described in the previous section, a hash
table is needed to match new keys with previously encountered keys.
But there is no guarantee that the hash table itself will not grow
arbitrarily large for data sets with many unique keys.  We need an
algorithm for generating KMV pairs that operates within a small
number of memory pages and performs a minimal number of passes through
the KV disk file.  An algorithm that meets these goals is diagrammed
in Figure \ref{fig:collate}; it reads the KV pages at most four times, and
writes out new KV or KMV pages at most three times.  It also uses a
finite-size in-memory hash table, that stores unique keys as they are
encountered while looping over the KV pages, as well as auxiliary
information needed to construct the output pages of KMV pairs.

\begin{figure}
\includegraphics[width=\textwidth,angle=-90]{fig_collate2.pdf}
\caption{Multi-pass algorithm for converting KV data to KMV data.  The
vertical lines represent out-of-core data sets.  KV is key/value
pairs; HT is an in-memory hash table; KMV is key/multivalue pairs.}
\label{fig:collate}
\end{figure}

@@

In summary, this data reorganization for the {\it collate()} operation
is somewhat complex, but requires only a small, constant number of
passes through the data.  The KV datums are read from disk at most four
times, and written to disk three times.  The first two write passes
reorganize KV pairs into partition and set files; the final write pass
creates the KMV data set.  Depending on the size and characteristics
of the KV datums (e.g., the number of unique keys), some of
these passes may not be needed.  By contrast, a full out-of-core sort
of the KV pairs, performed via a merge sort as discussed below,
requires $O(\log{M})$ read/write passes through the KV data, where $M$
is the number of pages of KV pairs, which can be a large number for
big data sets.

The memory page requirements for the above algorithm are as follows.
Two contiguous pages of memory are used for the hash table.
Intermediate passes 2 and 3 can require numerous partition or set
files to be opened simultaneously and written to.  To avoid small
writes of individual KV datums, a buffer of at least 16K bytes is
allocated for each file.  The precise number of simultaneously open
files is difficult to bound, but typically one or two additional pages of 
memory suffice for all needed buffers.  Thus the total number of pages required
for the data reorganization stage of the {\it collate()} operation is
less than the seven used by the all-to-all communication stage.

Other MR-MPI library calls discussed in the previous section, such as
{\it clone()}, {\it collapse()}, and {\it gather()},
can also operate in out-of-core mode, typically with one pass
through the KV or KMV data and the use of one or two memory pages.
Sorting KV pairs, by key or value, is an exception.  An out-of-core
merge sort is performed as follows.  Two pages of KV datums are read
from disk.  Each is sorted using the C-library quicksort() function,
as discussed in the previous section.  The two pages are then scanned
and merged into a new file which is written to disk as its associated
memory page fills up.  This process requires five memory pages, which
includes vectors of KV datum pointers and lengths needed by the
in-memory quicksort() operation.  Once all pairs of pages have been
merged, the sort continues in a recursive fashion, merging pairs of
files into a new third file, without the need to perform additional
in-memory sorts.  Thus the overall memory requirement is five pages.  The
number of read/write passes through the KV data set for the merge sort
is $O(\log{M})$, where $M$ is the number of pages of KV pairs.

>>>>

When a user program creates a MapReduce object, a ``pagesize'' can be
specified, which defaults to 64 Mbytes.  As described below, each
MR-MPI operation is constrained to use no more than a handful of these
pages.  The {\it pagesize} setting can be as small as 1 Mbyte or as
large as desired, though the user should ensure the allocated pages
fit in physical memory; otherwise a processor may allocate slow
virtual memory.  The {\it pagesize} is also the typical size of
individual reads and writes to the temporary disk files; hence a
reasonable {\it pagesize} ensures good I/O performance.

@@

When the communication stage of the {\it collate()} operation is
complete, each processor has a set of KV pages, stored in an
out-of-core KV file, that need to be reorganized into a set of KMV
pages, likewise stored in a new out-of-core KMV file.  In principle,
creating one page of KMV pairs would require all KV pages be scanned
to find all the keys that contribute values to that page.  Doing this
for each output page of KMV pairs could be prohibitively expensive.  A
related issue is that, as described in the previous section, a hash
table is needed to match new keys with previously encountered keys.
But there is no guarantee that the hash table itself will not grow
arbitrarily large for data sets with many unique keys.  We need an
algorithm for generating KMV pairs that operates within a small number
of memory pages and performs a minimal number of passes through the KV
disk file.  An algorithm that meets these goals is diagrammed in
Figure \ref{fig:collate}; it reads the KV pages at most four times,
and writes out new KV or KMV pages at most three times.  It also uses
a finite-size in-memory hash table, that stores unique keys as they
are encountered while looping over the KV pages, as well as auxiliary
information needed to construct the output pages of KMV pairs.

\begin{figure}
\includegraphics[width=\textwidth,angle=-90]{fig_collate2.pdf}
\caption{\it Multi-pass algorithm for converting KV data to KMV data.
The vertical lines represent out-of-core data sets.  KV is key/value
pairs; HT is an in-memory hash table; KMV is key/multivalue pairs.}
\label{fig:collate}
\end{figure}

@@

In summary, this data reorganization for the {\it collate()} operation
is somewhat complex, but requires only a small, constant number of
passes through the data.  The KV datums are read from disk at most
four times, and written to disk three times.  The first two write
passes reorganize KV pairs into partition and set files; the final
write pass creates the KMV data set.  Depending on the size and
characteristics of the KV datums (e.g., the number of unique keys),
some of these passes may not be needed.  By contrast, a full
out-of-core sort of the KV pairs, performed via a merge sort as
discussed below is still an $O(N\log_2{N})$ operation.  The number of
read/write passes through the KV data files depends on the amount of
KV data that can be held in memory, but can be a large number for big
data sets.

The memory page requirements for the out-of-core {\it collate()}
operation are as follows.  Two contiguous pages of memory are used for
the hash table.  Intermediate passes 2 and 3 can require numerous
partition or set files to be opened simultaneously and written to.  To
avoid small writes of individual KV datums, a buffer of at least 16K
bytes is allocated for each file.  The precise number of
simultaneously open files is difficult to bound, but typically one or
two additional pages of memory suffice for all needed buffers.  Thus
the total number of pages required for the data reorganization stage
of the {\it collate()} operation is less than the seven used by the
all-to-all communication stage.

Other MR-MPI library calls discussed in the previous section, such as
{\it clone()}, {\it collapse()}, and {\it gather()}, can also operate
in out-of-core mode, typically with one pass through the KV or KMV
data and the use of one or two memory pages.  Sorting KV pairs, by key
or value, is an exception.  An out-of-core merge sort is performed as
follows.  Two pages of KV datums are read from disk.  Each is sorted
using the C-library quicksort() function, as discussed in the previous
section.  The two pages are then scanned and merged into a new file
which is written to disk as its associated memory page fills up.  This
process requires five memory pages, which includes vectors of KV datum
pointers and lengths needed by the in-memory quicksort() operation.
Once all pairs of pages have been merged, the sort continues in a
recursive fashion, merging pairs of files into a new third file,
without the need to perform additional in-memory sorts.  Thus the
overall memory requirement is five pages.  The number of read/write
passes through the KV data set for the merge sort is $O(\log_2{M})$,
where $M$ is the number of pages of KV pairs.  The number of passes
could be reduced at the cost of allocating more in-memory pages.
